A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-08 paper.tex
Update

\section*{Event vs Genome ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

@@

\begin{figure}
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (3, 1) {};
\node (s4) [greynode] at (1, 2) {};
\node (s5) [greynode] at (5, 3) {};
\node (s6) [greynode] at (3, 4) {};

% \node [anchor=north west] at (0,6) {A};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

% \node [anchor=north west] at (9,6) {B};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

% \node [anchor=north west] at (9,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG and a concrete
encoding of the information. Node \noderef{d} is a recombination
event as it is associated with a non-null breakpoint, and has
two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
common ancestor events.
}
\end{figure}

@@

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As demonstrated in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

@@

% Formally, we can define a gARG as a set of edges $E$, where each
% element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
% $c$ is the child node, $p$ is the parent node and $I$ is the set of
% disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
% also phrase this in terms of the ancestral material of a sample---see
% the next section).
% Deriving the local tree at a point $x$
% follows largely the same pattern as for an eARG but is somewhat
% simpler because coordinate information is directly associated with
% edges. Starting with the sample nodes $S$ we traverse
% rootwards through the graph from child to parent. At each node $u$, we find an
% edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
% and move to the next node $p$.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
might seem of little practical importance.
However, there are substantial practical benefits
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computationally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.

>>>>

% \section*{Genome ARGs}\label{eARG}
% In keeping with its original derivation in terms of a stochastic
% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
% ARG data structure is
% in terms of events.
% Nodes represent
% common ancestor or recombination events that occurred in the
% history of some set of sample genomes, and edges represent ancestral
% lineages~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage, and in a recombination
% event a lineage is split into two independent
% ancestral lineages.
% % The final vital
% % detail is that we associate a breakpoint with each recombination
% % event.
% % This information is sufficient to uniquely
% % define the local genealogical trees at every position along the genome,
% % which is the basic requirement for a data structure encoding
% % genome-wide genetic ancestry.
% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
% is usually concretely encoded by associating a recombination breakpoint
% (which is null for common ancestor events) and an ordered list of parents
% with each node. We refer to such a  data structure as an
% ``event ARG'' (eARG).
% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.

@@

% \begin{figure}
% \centering
% % \begin{tabular}{cc}
% \begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
% \tikzset{greynode/.style={circle,fill,inner sep=1},
% nodelabel/.style={font=\footnotesize}}

% \node (s0) [greynode] at (0, 0) {};
% \node (s1) [greynode] at (3, 0) {};
% \node (s2) [greynode] at (6, 0) {};
% \node (s3) [greynode] at (3, 1) {};
% \node (s4) [greynode] at (1, 2) {};
% \node (s5) [greynode] at (5, 3) {};
% \node (s6) [greynode] at (3, 4) {};

% % \node [anchor=north west] at (0,6) {A};
% \node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
% \foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
% \foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
% \foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
% \foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

% %% Edges
% \draw (s1) -- (s3);
% \draw (s0) |- (s4);
% \draw (s4) -- (2,2) |- (s3);
% \draw (s4) |- (s6);
% \draw (s3) -- (4,1) |- (s5);
% \draw (s2) |- (s5);
% \draw (s5) |- (s6);

% % \node [anchor=north west] at (9,6) {B};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{c|c|l}
% % \multicolumn{2}{c}{Breakpoints}\\
% Node & Breakpoint & Parents\\
% \hline
% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{g}$ & $\varnothing$ & []\\
% \end{tabular}};

% % \node [anchor=north west] at (9,6) {C};
% % \node [nodelabel,anchor=north west] at ($(10,5)$) {
% % \begin{tabular}{ll}
% % Edges & Intervals\\
% % \hline
% % $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% % $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% % $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% % $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% % $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% % $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% % $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% % \end{tabular}};

% \end{tikzpicture}
% \caption{\label{fig-arg-data-structure}
% A graphical depiction of a classical event ARG and a concrete
% encoding of the information. Node \noderef{d} is a recombination
% event as it is associated with a non-null breakpoint, and has
% two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
% common ancestor events.
% }
% \end{figure}

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
to avoid confusion, we refer to as a ``genome ARG``, or gARG.
(We contrast this with the classical definition derived from the
coalescent process in later sections.)
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure)
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

@@

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure; see
Section XXX),
we traverse the graph rootwards from the leaves.
Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$.

[COMMENT ON FIG 1C, and unary nodes in the trees]

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
% The difference between eARGs and gARGs are subtle, and whether
% we choose to interpret nodes as events or genomes
% might seem of little practical importance.
% However, there are substantial practical benefits
% from using the gARG encoding, allowing differential
% levels of information about the recombination process to be
% represented in a computationally efficient way.
% The key to this is not so much in how we interpret nodes
% (see Appendix XXX), but specifically in how
% we represent the process of genetic inheritance.
% The critical distinction is that
% the gARG encoding allows us to directly describe
% genetic inheritance in multiple non-contiguous genome intervals
% between an ancestor and a descendant, something that
% cannot be done by storing recombination breakpoints.
